---
title: "Which way of combining forecasts holds up out of sample?"
subtitle: "Aggregation rules on GJP ACE (2011–2015) and the ForecastBench 2024 human round"
format:
  html:
    embed-resources: true
    toc: true
    fig-width: 8
    fig-height: 4.5
execute:
  echo: false
  warning: false
---

```{r setup}
for (f in c("ingest.R", "scoring.R", "plotting.R", "aggregate.R",
            "experiments.R")) {
  source(file.path("..", "R", f))
}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
con <- DBI::dbConnect(
  duckdb::duckdb(),
  dbdir = file.path("..", "data", "db", "gjp.duckdb"), read_only = TRUE
)
q <- function(sql) DBI::dbGetQuery(con, sql)

settings <- as.list(stats::setNames(
  q("SELECT value FROM agg_settings")$value,
  q("SELECT parameter FROM agg_settings")$parameter
))
results <- q("SELECT * FROM agg_results")
sweep <- q("SELECT * FROM agg_sweep")
gjp_q <- q("SELECT * FROM agg_gjp_questions")
fb_q <- q("SELECT * FROM agg_fb_questions")
curve <- q("SELECT * FROM agg_crowd_size")

rule_label <- function(rule) {
  lab <- c(
    mean = "mean",
    median = "median",
    trimmed_mean = sprintf("trimmed mean (%.0f%%)", 100 * settings$trim),
    hd_trim = sprintf("HD-trim (%.0f%%)", 100 * settings$hd_trim),
    soften_mean = sprintf("softened mean (%.0f%%)", 100 * settings$soften),
    geo_mean_odds = "geometric mean of odds",
    extremized = sprintf("extremized (a = %.2g)", settings$extremize_a),
    neyman = "Neyman extremized",
    "individual forecaster" = "individual forecaster"
  )
  lab[rule] <- ifelse(is.na(lab[rule]), rule, lab[rule])
  out <- unname(lab[rule])
  sel <- grepl("^select_crowd_k", rule)
  out[sel] <- sprintf("select crowd (top %s)",
                      sub("^select_crowd_k", "", rule[sel]))
  out
}

cohort_label <- c(
  "independent individuals (yrs 3-4)" = "GJP independents, years 3-4",
  fb_public = "ForecastBench public",
  fb_super = "ForecastBench supers"
)
results$panel <- factor(unname(cohort_label[results$cohort]),
                        levels = unname(cohort_label))
results$label <- rule_label(results$rule)
gjp <- results[results$corpus == "gjp", ]
crowd_only <- gjp[gjp$rule != "individual forecaster", ]
best_gjp <- crowd_only[which.min(crowd_only$mean_brier), ]
worst_gjp <- crowd_only[which.max(crowd_only$mean_brier), ]
ind_gjp <- gjp[gjp$rule == "individual forecaster", ]

# The value of pooling at all against the value of pooling well: the step
# from the average member to a plain mean, over the step from a plain mean to
# the best rule. Anchored on the mean rather than on the weakest rule in the
# field, so it does not depend on which rules were entered.
step_ratio <- function(coh) {
  df <- results[results$cohort == coh, ]
  ind <- df$mean_brier[df$rule == "individual forecaster"]
  plain <- df$mean_brier[df$rule == "mean"]
  best <- min(df$mean_brier[df$rule != "individual forecaster"])
  (ind - plain) / (plain - best)
}
plain_mean <- function(coh) {
  results$mean_brier[results$cohort == coh & results$rule == "mean"]
}
```

## Scope

The calibration note measured how good these forecasters are one at a time.
This note asks the other half of the question: pooled into a crowd, how much
better do they get, and which pooling rule survives being tuned on one part
of the data and applied to another.

Nine aggregation rules run against the same forecasts. Three are taken from
`aggutils`, the reference implementation of the trimming and extremizing
methods published in Powell et al. (2024) and Neyman & Roughgarden (2022);
the other six are implemented here, with unit tests that check the two agree
wherever they overlap. The softened mean is among the six for a specific
reason: `aggutils::soften_mean` takes its input on the 0-100 scale but
compares the mean against 0.5 rather than 50, so it trims the upper tail of
any crowd averaging above half a percent, including crowds that lean low.
What runs here is the rule as the package documents it, with that threshold
corrected and nothing else changed.

## Protocol

**The crowd is scored exactly as its members were.** For GJP that means a day
panel: on each day of a question's scoring window, every independent
forecaster's standing forecast is collected, the rule turns them into one
probability, and that probability is scored. A question's score is the mean
over its days; questions then average unweighted. The panel holds
`r format(q("SELECT count(*) FROM gjp_panel")[[1]], big.mark = ",")`
forecaster-days over
`r format(q("SELECT count(DISTINCT ifp_id || day) FROM gjp_panel")[[1]], big.mark = ",")`
question-days. For ForecastBench, where a forecaster answers once per
horizon, the crowd is pooled per question-horizon.

**Tuning is separated from evaluation by construction.** Every free
parameter — trim fractions, the extremizing constant, the select-crowd size —
is chosen on GJP years 1–2 and then frozen. Results below are GJP years 3–4,
never seen during tuning, plus the ForecastBench 2024 round as external
validation. The split is enforced in code rather than by care: the tuner
raises an error if it is handed scores carrying an evaluation year, and a
test exercises that.

**Select-crowd cannot see its own future.** A forecaster's trailing accuracy
on a given day is computed only from questions that finished scoring
*strictly before* that day, joined as-of the forecast date, and a forecaster
needs `r select_min_history()` finished questions before being eligible.
Where too few forecasters qualify the rule falls back to the whole-crowd
median and the fallback days are counted — on the evaluation years the
fallback rate was
`r sprintf("%.1f%%", 100 * mean(gjp_q$fallback_rate[grepl("select", gjp_q$rule)]))`.
Select-crowd is GJP-only; a single ForecastBench round carries no history to
select on.

## Tuning

```{r tuning}
sweep$family <- sub("[@0-9.]*$", "", sub("_k[0-9]+$", "_k", sweep$rule))
sweep$value <- suppressWarnings(as.numeric(
  ifelse(grepl("@", sweep$rule), sub("^.*@", "", sweep$rule),
         sub("^select_crowd_k", "", sweep$rule))
))
chosen <- data.frame(
  family = c("trimmed_mean", "hd_trim", "soften_mean", "extremized",
             "select_crowd_k"),
  value = c(settings$trim, settings$hd_trim, settings$soften,
            settings$extremize_a, settings$select_k)
)
tune_tbl <- merge(chosen, sweep, by = c("family", "value"))
tune_tbl$grid <- vapply(tune_tbl$family, function(f) {
  paste(sort(sweep$value[sweep$family == f]), collapse = ", ")
}, character(1))
tune_tbl$frozen <- format(tune_tbl$value, trim = TRUE)
knitr::kable(
  tune_tbl[, c("family", "grid", "frozen", "brier")],
  digits = 3,
  col.names = c("family", "grid searched", "frozen value",
                "tuning-year Brier"),
  caption = "Parameters chosen on GJP years 1-2. Every frozen value is an
             interior minimum of its grid, not a boundary pick."
)
```

Extremizing has a clear optimum near
`r sprintf("a = %.2g", settings$extremize_a)`: pooling in log odds and then
pushing the result away from 50% helps, but only up to a point, and past it
the crowd becomes overconfident. The trimming families keep improving until
they are discarding roughly `r sprintf("%.0f%%", 100 * settings$trim)` of the
crowd from each side, which is close to behaving like a median — consistent
with trimming's usual role of removing the confident tails rather than
reshaping the middle.

## The horse race

```{r horse-race, fig.height=5.2}
plot_df <- results
plot_df$is_baseline <- plot_df$rule == "individual forecaster"
ord <- gjp$label[order(gjp$mean_brier, decreasing = TRUE)]
plot_df$label <- factor(plot_df$label,
                        levels = c(setdiff(ord, "individual forecaster"),
                                   "individual forecaster"))
p_race <- ggplot(plot_df, aes(mean_brier, label, color = is_baseline)) +
  geom_linerange(aes(xmin = ci_low, xmax = ci_high), linewidth = 0.6,
                 alpha = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~panel, scales = "free_x") +
  scale_color_manual(values = c(`FALSE` = cfe_palette[1],
                                `TRUE` = cfe_palette[2]),
                     labels = c("aggregation rule", "average member")) +
  labs(x = "mean question Brier [0-2], lower is better", y = NULL,
       title = "Aggregation rules, out of sample",
       subtitle = paste(
         "Parameters frozen on GJP years 1-2; bars are 95% question-cluster",
         "bootstrap intervals.\nSelect crowd runs on GJP only; the three",
         "panels are different question sets."
       )) +
  theme_cfe()
save_fig(p_race, "horse-race.png", h = 5.2)
```

```{r results-table}
tbl <- results[order(results$corpus, results$cohort, results$mean_brier), ]
knitr::kable(
  data.frame(
    corpus = tbl$corpus, cohort = tbl$cohort, rule = tbl$label,
    questions = tbl$n_questions, scored = tbl$n_scored,
    brier = tbl$mean_brier, low = tbl$ci_low, high = tbl$ci_high
  ),
  digits = 3,
  col.names = c("corpus", "cohort", "rule", "questions", "scored rows",
                "mean Brier [0-2]", "2.5%", "97.5%")
)
```

## What the crowd is worth

```{r crowd-value}
ind_q <- individual_baseline_gjp(con, evaluation_years())
best_q <- gjp_q[gjp_q$rule == best_gjp$rule, c("question_id", "brier")]
paired <- merge(ind_q, best_q, by = "question_id",
                suffixes = c("_ind", "_crowd"))
paired$d <- paired$brier_ind - paired$brier_crowd
b_gjp <- bootstrap_questions(paired, function(df) mean(df$d))

fb_ind <- individual_baseline_fb(con)
fb_best <- do.call(rbind, lapply(split(fb_q, fb_q$cohort), function(df) {
  m <- stats::aggregate(brier ~ rule, data = df, FUN = mean)
  keep <- m$rule[which.min(m$brier)]
  df[df$rule == keep, ]
}))
fb_pair <- merge(
  fb_ind, fb_best,
  by = c("cohort", "question_id", "resolution_date"),
  suffixes = c("_ind", "_crowd")
)
fb_pair$d <- fb_pair$brier_ind - fb_pair$brier_crowd
fb_boot <- lapply(split(fb_pair, fb_pair$cohort), function(df) {
  c(rule = df$rule[1], bootstrap_questions(df, function(x) mean(x$d)))
})

gain <- data.frame(
  comparison = c(
    sprintf("GJP years 3-4: average member - %s", best_gjp$label),
    sprintf("ForecastBench public: average member - %s",
            rule_label(fb_boot$fb_public[["rule"]])),
    sprintf("ForecastBench supers: average member - %s",
            rule_label(fb_boot$fb_super[["rule"]]))
  ),
  estimate = c(b_gjp["estimate"],
               as.numeric(fb_boot$fb_public["estimate"]),
               as.numeric(fb_boot$fb_super["estimate"])),
  low = c(b_gjp["low"], as.numeric(fb_boot$fb_public["low"]),
          as.numeric(fb_boot$fb_super["low"])),
  high = c(b_gjp["high"], as.numeric(fb_boot$fb_public["high"]),
           as.numeric(fb_boot$fb_super["high"]))
)
knitr::kable(
  gain, digits = 3, row.names = FALSE,
  col.names = c("paired difference in mean Brier (positive = crowd better)",
                "estimate", "2.5%", "97.5%"),
  caption = "Question-cluster bootstrap on paired per-question differences,
             2,000 replicates."
)
```

On the GJP evaluation years the best rule scores
`r sprintf("%.3f", best_gjp$mean_brier)` against
`r sprintf("%.3f", ind_gjp$mean_brier)` for the average member of the same
crowd on the same days — a reduction of
`r sprintf("%.0f%%", 100 * (1 - best_gjp$mean_brier / ind_gjp$mean_brier))`,
with the paired interval well clear of zero. That is the wisdom-of-crowds
effect at full strength, and most of it needs no cleverness: the plain mean,
the most obvious rule available, already takes the score to
`r sprintf("%.3f", plain_mean(best_gjp$cohort))`, and going from there to the
best rule is worth a further
`r sprintf("%.3f", plain_mean(best_gjp$cohort) - best_gjp$mean_brier)`.
Pooling at all is the larger step, by a factor of
`r sprintf("%.1f", step_ratio(best_gjp$cohort))` — not a different order of
magnitude. The choice of rule earns its keep; it just earns less than the
decision to pool.

## Crowd size

```{r crowd-size, fig.height=4.2}
show <- c("mean", "median", "extremized")
curve_plot <- curve[curve$rule %in% show, ]
curve_plot$label <- rule_label(curve_plot$rule)
ends <- curve_plot[curve_plot$size == max(curve_plot$size), ]
p_size <- ggplot(curve_plot, aes(size, brier, color = label)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  geom_text(data = ends, aes(label = label), hjust = -0.12, vjust = 0.4,
            size = 3.1, show.legend = FALSE) +
  scale_x_log10(breaks = sort(unique(curve$size)),
                expand = expansion(mult = c(0.04, 0.22))) +
  scale_color_manual(values = cfe_palette[1:3], guide = "none") +
  labs(x = "forecasters drawn at random (log scale)",
       y = "mean Brier [0-2]",
       title = "Accuracy against crowd size",
       subtitle = paste(
         "GJP years 3-4; on each scoring day, up to n of that day's standing",
         "forecasters are drawn,\nso every size is scored on the same days.",
         "5 draws per question and size, up to 30 days each."
       )) +
  theme_cfe()
save_fig(p_size, "crowd-size.png", h = 4.2)
```

```{r crowd-size-table}
wide <- stats::reshape(
  transform(curve, label = rule_label(rule))[, c("label", "size", "brier")],
  idvar = "label", timevar = "size", direction = "wide"
)
names(wide) <- sub("^brier\\.", "n = ", names(wide))
knitr::kable(wide, digits = 3, row.names = FALSE,
             caption = "Mean Brier by crowd size, including HD-trim.")
```

Two things separate the rules. The simple mean flattens early: past roughly
25 forecasters it stops improving, because averaging probabilities pulls the
aggregate towards the middle no matter how many people agree. Extremizing
keeps gaining, because a larger crowd both cancels more noise and justifies
more confidence in the pooled signal. At one forecaster the extremized rule
is the worst of the three, which is the same fact seen from the other side:
there is nothing to be confident about in a single opinion.

At n = 1 the mean and the median are just the drawn forecast, and they land on
`r sprintf("%.3f", curve$brier[curve$rule == "mean" & curve$size == 1])`
against `r sprintf("%.3f", ind_gjp$mean_brier)` for the average member in the
table above, a gap of
`r sprintf("%.3f", abs(curve$brier[curve$rule == "mean" & curve$size == 1] - ind_gjp$mean_brier))`
left over from thinning the days. Agreement there is the check that the curve
and the table are measuring the same thing. That anchoring is the point of drawing the
subcrowd from each day's standing forecasters
rather than once per question: a crowd sampled once per question covers more
of a question's hard early stretch as it grows, and the extra coverage would
otherwise show up as a size effect.

## Transfer to the modern round

```{r transfer}
rank_in <- function(coh, rule) {
  df <- results[results$cohort == coh & results$rule != "individual forecaster", ]
  which(df$rule[order(df$mean_brier)] == rule)
}
best_in <- function(coh) {
  df <- results[results$cohort == coh & results$rule != "individual forecaster", ]
  df$label[which.min(df$mean_brier)]
}
n_rules <- sum(results$cohort == "fb_public" &
                 results$rule != "individual forecaster")
transfer <- results[results$corpus == "forecastbench" &
                      results$rule %in% c("geo_mean_odds", "extremized",
                                          "neyman", "median",
                                          "trimmed_mean", "mean"), ]
knitr::kable(
  data.frame(cohort = transfer$cohort, rule = transfer$label,
             brier = transfer$mean_brier)[
               order(transfer$cohort, transfer$mean_brier), ],
  digits = 3, row.names = FALSE,
  col.names = c("cohort", "rule", "mean Brier [0-2]")
)
```

The extremizing constant tuned on GJP does not travel. On ForecastBench
superforecasters it is close to the front, and near-identical to untuned
pooling in log odds. On the public cohort it finishes
`r rank_in("fb_public", "extremized")`th of `r n_rules`, behind even the plain
mean, while `r tolower(best_in("fb_public"))` wins.

This is the calibration note's finding arriving from the other direction.
The GJP independents are collectively underconfident, so pushing their pooled
forecast outwards helps. The ForecastBench public cohort already flattens
above 50% — high probabilities not matched by outcomes — and extremizing an
overconfident crowd makes it worse. Extremizing constants are properties of a
forecaster pool, not of aggregation in general, and a constant carried across
pools without checking its calibration is a liability. Neyman's rule takes its
factor from crowd size rather than from a tuned constant, which spares it the
transfer question but not the underlying one: it still extremizes, so while it
is the best rule on the supers it finishes
`r rank_in("fb_public", "neyman")`th of `r n_rules` on the public — one place
ahead of the tuned constant, not in a different class.

## Reconciliation

The ordering matches what the aggregation literature reports: pooling in log
odds beats pooling probabilities in all three cohorts, trimming beats the raw
mean in all three, and extremized and skill-selected crowds sit at the top on
the corpus they were tuned on (Satopää et al. 2014; Mellers et al. 2014;
Powell et al. 2024; Neyman & Roughgarden 2022). Select-crowd sits ahead of
every non-extremizing rule and behind both extremizing ones, which is what to
expect at this crowd size: with `r sprintf("%s", settings$select_k)`
forecasters kept out of hundreds, selecting on skill buys less than
extremizing buys from the whole crowd's cancelled noise.

How far pooling outruns rule choice depends on the pool. Measured as the step
from the average member to a plain mean, over the step from that mean to the
best rule, it is `r sprintf("%.1f", step_ratio(best_gjp$cohort))` on the GJP
independents, `r sprintf("%.1f", step_ratio("fb_public"))` on the
ForecastBench public crowd, and
`r sprintf("%.1f", step_ratio("fb_super"))` on the supers. Pooling is the
bigger lever in all three, and by the widest margin on the public crowd,
whose members are furthest behind their own aggregate. The supers are the
narrow case: the whole field of rules spans
`r sprintf("%.3f", diff(range(results$mean_brier[results$cohort == "fb_super" & results$rule != "individual forecaster"])))`
and every gap in it is smaller than its own interval. Aggregation buys the
most where the individuals disagree the most, which is the mechanism, not an
incidental finding.

Scores here are raw Brier throughout, on the two-option scale. The
ForecastBench leaderboard's published figures are difficulty-adjusted and
rescaled, so they are a different statistic; the calibration note documents
that difference and reconciles the raw reproduction against it.

## Reproducibility

`make experiments` rebuilds the day panel, re-runs the tuning sweep, and
writes every table below to `analysis/results/`. `make analyze` renders this
document.

```{r teardown}
DBI::dbDisconnect(con, shutdown = TRUE)
```
