---
title: "Which way of combining forecasts holds up out of sample?"
subtitle: "Aggregation rules on GJP ACE (2011–2015) and the ForecastBench 2024 human round"
format:
  html:
    embed-resources: true
    toc: true
    fig-width: 8
    fig-height: 4.5
execute:
  echo: false
  warning: false
---

```{r setup}
for (f in c("ingest.R", "scoring.R", "plotting.R", "aggregate.R",
            "experiments.R")) {
  source(file.path("..", "R", f))
}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
con <- DBI::dbConnect(
  duckdb::duckdb(),
  dbdir = file.path("..", "data", "db", "gjp.duckdb"), read_only = TRUE
)
q <- function(sql) DBI::dbGetQuery(con, sql)

settings <- as.list(stats::setNames(
  q("SELECT value FROM agg_settings")$value,
  q("SELECT parameter FROM agg_settings")$parameter
))
results <- q("SELECT * FROM agg_results")
sweep <- q("SELECT * FROM agg_sweep")
gjp_q <- q("SELECT * FROM agg_gjp_questions")
fb_q <- q("SELECT * FROM agg_fb_questions")
curve <- q("SELECT * FROM agg_crowd_size")

rule_label <- function(rule) {
  lab <- c(
    mean = "mean",
    median = "median",
    trimmed_mean = sprintf("trimmed mean (%.0f%%)", 100 * settings$trim),
    hd_trim = sprintf("HD-trim (%.0f%%)", 100 * settings$hd_trim),
    soften_mean = sprintf("softened mean (%.0f%%)", 100 * settings$soften),
    geo_mean_odds = "geometric mean of odds",
    extremized = sprintf("extremized (a = %.2g)", settings$extremize_a),
    neyman = "Neyman extremized",
    "individual forecaster" = "individual forecaster"
  )
  lab[rule] <- ifelse(is.na(lab[rule]), rule, lab[rule])
  out <- unname(lab[rule])
  sel <- grepl("^select_crowd_k", rule)
  out[sel] <- sprintf("select crowd (top %s)",
                      sub("^select_crowd_k", "", rule[sel]))
  out
}

cohort_label <- c(
  "independent individuals (yrs 3-4)" = "GJP independents, years 3-4",
  fb_public = "ForecastBench public",
  fb_super = "ForecastBench supers"
)
results$panel <- factor(unname(cohort_label[results$cohort]),
                        levels = unname(cohort_label))
results$label <- rule_label(results$rule)
gjp <- results[results$corpus == "gjp", ]
crowd_only <- gjp[gjp$rule != "individual forecaster", ]
best_gjp <- crowd_only[which.min(crowd_only$mean_brier), ]
ind_gjp <- gjp[gjp$rule == "individual forecaster", ]
```

## Scope

The calibration note measured how good these forecasters are one at a time.
This note asks the other half of the question: pooled into a crowd, how much
better do they get, and which pooling rule survives being tuned on one part
of the data and applied to another.

Nine aggregation rules run against the same forecasts. Four are taken from
`aggutils`, the reference implementation of the trimming and extremizing
methods published in Powell et al. (2022) and Neyman & Roughgarden (2021);
the rest are implemented here, with unit tests that check the two agree
wherever they overlap.

## Protocol

**The crowd is scored exactly as its members were.** For GJP that means a day
panel: on each day of a question's scoring window, every independent
forecaster's standing forecast is collected, the rule turns them into one
probability, and that probability is scored. A question's score is the mean
over its days; questions then average unweighted. The panel holds
`r format(q("SELECT count(*) FROM gjp_panel")[[1]], big.mark = ",")`
forecaster-days over
`r format(q("SELECT count(DISTINCT ifp_id || day) FROM gjp_panel")[[1]], big.mark = ",")`
question-days. For ForecastBench, where a forecaster answers once per
horizon, the crowd is pooled per question-horizon.

**Tuning is separated from evaluation by construction.** Every free
parameter — trim fractions, the extremizing constant, the select-crowd size —
is chosen on GJP years 1–2 and then frozen. Results below are GJP years 3–4,
never seen during tuning, plus the ForecastBench 2024 round as external
validation. The split is enforced in code rather than by care: the tuner
raises an error if it is handed scores carrying an evaluation year, and a
test exercises that.

**Select-crowd cannot see its own future.** A forecaster's trailing accuracy
on a given day is computed only from questions that finished scoring
*strictly before* that day, joined as-of the forecast date, and a forecaster
needs `r select_min_history()` finished questions before being eligible.
Where too few forecasters qualify the rule falls back to the whole-crowd
median and the fallback days are counted — on the evaluation years the
fallback rate was
`r sprintf("%.1f%%", 100 * mean(gjp_q$fallback_rate[grepl("select", gjp_q$rule)]))`.
Select-crowd is GJP-only; a single ForecastBench round carries no history to
select on.

## Tuning

```{r tuning}
sweep$family <- sub("[@0-9.]*$", "", sub("_k[0-9]+$", "_k", sweep$rule))
sweep$value <- suppressWarnings(as.numeric(
  ifelse(grepl("@", sweep$rule), sub("^.*@", "", sweep$rule),
         sub("^select_crowd_k", "", sweep$rule))
))
chosen <- data.frame(
  family = c("trimmed_mean", "hd_trim", "soften_mean", "extremized",
             "select_crowd_k"),
  value = c(settings$trim, settings$hd_trim, settings$soften,
            settings$extremize_a, settings$select_k)
)
tune_tbl <- merge(chosen, sweep, by = c("family", "value"))
tune_tbl$grid <- vapply(tune_tbl$family, function(f) {
  paste(sort(sweep$value[sweep$family == f]), collapse = ", ")
}, character(1))
tune_tbl$frozen <- format(tune_tbl$value, trim = TRUE)
knitr::kable(
  tune_tbl[, c("family", "grid", "frozen", "brier")],
  digits = 3,
  col.names = c("family", "grid searched", "frozen value",
                "tuning-year Brier"),
  caption = "Parameters chosen on GJP years 1-2. Every frozen value is an
             interior minimum of its grid, not a boundary pick."
)
```

Extremizing has a clear optimum near
`r sprintf("a = %.2g", settings$extremize_a)`: pooling in log odds and then
pushing the result away from 50% helps, but only up to a point, and past it
the crowd becomes overconfident. The trimming families keep improving until
they are discarding roughly `r sprintf("%.0f%%", 100 * settings$trim)` of the
crowd from each side, which is close to behaving like a median — consistent
with trimming's usual role of removing the confident tails rather than
reshaping the middle.

## The horse race

```{r horse-race, fig.height=5.2}
plot_df <- results
plot_df$is_baseline <- plot_df$rule == "individual forecaster"
ord <- gjp$label[order(gjp$mean_brier, decreasing = TRUE)]
plot_df$label <- factor(plot_df$label,
                        levels = c(setdiff(ord, "individual forecaster"),
                                   "individual forecaster"))
p_race <- ggplot(plot_df, aes(mean_brier, label, color = is_baseline)) +
  geom_linerange(aes(xmin = ci_low, xmax = ci_high), linewidth = 0.6,
                 alpha = 0.8) +
  geom_point(size = 2) +
  facet_wrap(~panel, scales = "free_x") +
  scale_color_manual(values = c(`FALSE` = cfe_palette[1],
                                `TRUE` = cfe_palette[2]),
                     labels = c("aggregation rule", "average member")) +
  labs(x = "mean question Brier [0-2], lower is better", y = NULL,
       title = "Aggregation rules, out of sample",
       subtitle = paste(
         "Parameters frozen on GJP years 1-2; bars are 95% question-cluster",
         "bootstrap intervals.\nSelect crowd runs on GJP only; the three",
         "panels are different question sets."
       )) +
  theme_cfe()
save_fig(p_race, "horse-race.png", h = 5.2)
```

```{r results-table}
tbl <- results[order(results$corpus, results$cohort, results$mean_brier), ]
knitr::kable(
  data.frame(
    corpus = tbl$corpus, cohort = tbl$cohort, rule = tbl$label,
    questions = tbl$n_questions, scored = tbl$n_scored,
    brier = tbl$mean_brier, low = tbl$ci_low, high = tbl$ci_high
  ),
  digits = 3,
  col.names = c("corpus", "cohort", "rule", "questions", "scored rows",
                "mean Brier [0-2]", "2.5%", "97.5%")
)
```

## What the crowd is worth

```{r crowd-value}
ind_q <- individual_baseline_gjp(con, evaluation_years())
best_q <- gjp_q[gjp_q$rule == best_gjp$rule, c("question_id", "brier")]
paired <- merge(ind_q, best_q, by = "question_id",
                suffixes = c("_ind", "_crowd"))
paired$d <- paired$brier_ind - paired$brier_crowd
b_gjp <- bootstrap_questions(paired, function(df) mean(df$d))

fb_ind <- individual_baseline_fb(con)
fb_best <- do.call(rbind, lapply(split(fb_q, fb_q$cohort), function(df) {
  m <- stats::aggregate(brier ~ rule, data = df, FUN = mean)
  keep <- m$rule[which.min(m$brier)]
  df[df$rule == keep, ]
}))
fb_pair <- merge(
  fb_ind, fb_best,
  by = c("cohort", "question_id", "resolution_date"),
  suffixes = c("_ind", "_crowd")
)
fb_pair$d <- fb_pair$brier_ind - fb_pair$brier_crowd
fb_boot <- lapply(split(fb_pair, fb_pair$cohort), function(df) {
  c(rule = df$rule[1], bootstrap_questions(df, function(x) mean(x$d)))
})

gain <- data.frame(
  comparison = c(
    sprintf("GJP years 3-4: average member - %s", best_gjp$label),
    sprintf("ForecastBench public: average member - %s",
            rule_label(fb_boot$fb_public[["rule"]])),
    sprintf("ForecastBench supers: average member - %s",
            rule_label(fb_boot$fb_super[["rule"]]))
  ),
  estimate = c(b_gjp["estimate"],
               as.numeric(fb_boot$fb_public["estimate"]),
               as.numeric(fb_boot$fb_super["estimate"])),
  low = c(b_gjp["low"], as.numeric(fb_boot$fb_public["low"]),
          as.numeric(fb_boot$fb_super["low"])),
  high = c(b_gjp["high"], as.numeric(fb_boot$fb_public["high"]),
           as.numeric(fb_boot$fb_super["high"]))
)
knitr::kable(
  gain, digits = 3, row.names = FALSE,
  col.names = c("paired difference in mean Brier (positive = crowd better)",
                "estimate", "2.5%", "97.5%"),
  caption = "Question-cluster bootstrap on paired per-question differences,
             2,000 replicates."
)
```

On the GJP evaluation years the best rule scores
`r sprintf("%.3f", best_gjp$mean_brier)` against
`r sprintf("%.3f", ind_gjp$mean_brier)` for the average member of the same
crowd on the same days — a reduction of
`r sprintf("%.0f%%", 100 * (1 - best_gjp$mean_brier / ind_gjp$mean_brier))`,
with the paired interval well clear of zero. That is the wisdom-of-crowds
effect at full strength, and most of it is available from any sensible rule:
even the plain mean beats the average member by a wide margin. The choice of
rule is worth much less than the decision to aggregate at all — but it is not
nothing, and the ordering of rules is stable.

## Crowd size

```{r crowd-size, fig.height=4.2}
show <- c("mean", "median", "extremized")
curve_plot <- curve[curve$rule %in% show, ]
curve_plot$label <- rule_label(curve_plot$rule)
ends <- curve_plot[curve_plot$size == max(curve_plot$size), ]
p_size <- ggplot(curve_plot, aes(size, brier, color = label)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  geom_text(data = ends, aes(label = label), hjust = -0.12, vjust = 0.4,
            size = 3.1, show.legend = FALSE) +
  scale_x_log10(breaks = sort(unique(curve$size)),
                expand = expansion(mult = c(0.04, 0.22))) +
  scale_color_manual(values = cfe_palette[1:3], guide = "none") +
  labs(x = "forecasters drawn at random (log scale)",
       y = "mean Brier [0-2]",
       title = "Accuracy against crowd size",
       subtitle = paste(
         "GJP years 3-4; on each scoring day, up to n of that day's standing",
         "forecasters are drawn,\nso every size is scored on the same days.",
         "5 draws per question and size, up to 30 days each."
       )) +
  theme_cfe()
save_fig(p_size, "crowd-size.png", h = 4.2)
```

```{r crowd-size-table}
wide <- stats::reshape(
  transform(curve, label = rule_label(rule))[, c("label", "size", "brier")],
  idvar = "label", timevar = "size", direction = "wide"
)
names(wide) <- sub("^brier\\.", "n = ", names(wide))
knitr::kable(wide, digits = 3, row.names = FALSE,
             caption = "Mean Brier by crowd size, including HD-trim.")
```

Two things separate the rules. The simple mean flattens early: past roughly
25 forecasters it stops improving, because averaging probabilities pulls the
aggregate towards the middle no matter how many people agree. Extremizing
keeps gaining, because a larger crowd both cancels more noise and justifies
more confidence in the pooled signal. At one forecaster the extremized rule
is the worst of the three, which is the same fact seen from the other side:
there is nothing to be confident about in a single opinion.

At n = 1 the three rules coincide with the average member, which anchors the
curve to the baseline in the table above. Drawing the subcrowd from each
day's standing forecasters, rather than once per question, is what makes that
anchoring hold: a crowd sampled once per question covers more of a question's
hard early stretch as it grows, and the extra coverage would otherwise show
up as a size effect.

## Transfer to the modern round

```{r transfer}
rank_in <- function(coh, rule) {
  df <- results[results$cohort == coh & results$rule != "individual forecaster", ]
  which(df$rule[order(df$mean_brier)] == rule)
}
best_in <- function(coh) {
  df <- results[results$cohort == coh & results$rule != "individual forecaster", ]
  df$label[which.min(df$mean_brier)]
}
n_rules <- sum(results$cohort == "fb_public" &
                 results$rule != "individual forecaster")
transfer <- results[results$corpus == "forecastbench" &
                      results$rule %in% c("geo_mean_odds", "extremized",
                                          "neyman", "median",
                                          "trimmed_mean", "mean"), ]
knitr::kable(
  data.frame(cohort = transfer$cohort, rule = transfer$label,
             brier = transfer$mean_brier)[
               order(transfer$cohort, transfer$mean_brier), ],
  digits = 3, row.names = FALSE,
  col.names = c("cohort", "rule", "mean Brier [0-2]")
)
```

The extremizing constant tuned on GJP does not travel. On ForecastBench
superforecasters it is close to the front, and near-identical to untuned
pooling in log odds. On the public cohort it finishes
`r rank_in("fb_public", "extremized")`th of `r n_rules`, behind even the plain
mean, while `r tolower(best_in("fb_public"))` wins.

This is the calibration note's finding arriving from the other direction.
The GJP independents are collectively underconfident, so pushing their pooled
forecast outwards helps. The ForecastBench public cohort already flattens
above 50% — high probabilities not matched by outcomes — and extremizing an
overconfident crowd makes it worse. Extremizing constants are properties of a
forecaster pool, not of aggregation in general, and a constant carried across
pools without checking its calibration is a liability. Neyman's rule, which
derives its factor from crowd size rather than from a tuned constant, degrades
more gracefully: it is the best rule on the supers and mid-field on the
public.

## Reconciliation

The ordering matches what the aggregation literature reports: pooling in log
odds beats pooling probabilities, trimming beats the raw mean, extremized and
skill-selected crowds sit at the top on the corpus they were tuned on, and
the gain from aggregation dwarfs the gain from choosing among rules
(Satopää et al. 2014; Mellers et al. 2014; Powell et al. 2022;
Neyman & Roughgarden 2021). Select-crowd finishing near — not above — the
extremized aggregate is the expected result at this crowd size: with
`r sprintf("%s", settings$select_k)` forecasters selected out of hundreds,
the rule trades noise reduction for skill, and the two roughly cancel.

Scores here are raw Brier throughout, on the two-option scale. The
ForecastBench leaderboard's published figures are difficulty-adjusted and
rescaled, so they are a different statistic; the calibration note documents
that difference and reconciles the raw reproduction against it.

## Reproducibility

`make experiments` rebuilds the day panel, re-runs the tuning sweep, and
writes every table below to `analysis/results/`. `make analyze` renders this
document.

```{r teardown}
DBI::dbDisconnect(con, shutdown = TRUE)
```
