---
title: "Scoring and Aggregating Human Probability Forecasts on Two Public Corpora"
subtitle: "An out-of-sample comparison of nine aggregation rules"
author: "Alejandro Guerrero Padrés"
date: today
abstract: |
  Given many people's probability forecasts on the same question, how should
  they be scored, and how should they be combined? This paper answers both
  questions against two public corpora of real human forecasts: the Good
  Judgment Project's ACE tournament surveys (2011–2015, 3.1 million
  individual-level rows) and the ForecastBench 2024 human round. Individual
  accuracy and calibration are measured under precisely stated protocols, and
  nine aggregation rules are then compared out of sample, with every free
  parameter tuned on the first two tournament years and frozen before
  evaluation on the last two and on the modern round. Aggregating is worth far
  more than choosing how to aggregate: on the held-out years the best rule
  scores 0.204 against 0.422 for the average member of the same crowd on the
  same days, while the spread across rules is 0.086. Log-odds extremizing wins
  on the tournament data and keeps improving as the crowd grows, where the
  simple mean saturates near twenty-five forecasters. It does not travel: the
  same constant ranks seventh of eight on ForecastBench's public crowd, which
  is overconfident where the tournament's independents are underconfident.
  Extremizing constants are properties of a forecaster pool rather than of
  aggregation in general. The scored ForecastBench set reproduces the
  benchmark leaderboard's 521 question-horizons exactly, and the residual gap
  against its published scores is traced to the leaderboard's
  difficulty adjustment rather than left open.
format:
  pdf:
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    number-sections: true
    colorlinks: true
    fig-width: 6.5
    fig-height: 3.6
    fig-pos: "htbp"
    tbl-cap-location: bottom
bibliography: references.bib
link-citations: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
for (f in c("ingest.R", "scoring.R", "plotting.R", "aggregate.R",
            "experiments.R", "qa.R")) {
  source(file.path("..", "R", f))
}
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
con <- DBI::dbConnect(
  duckdb::duckdb(),
  dbdir = file.path("..", "data", "db", "gjp.duckdb"), read_only = TRUE
)
q <- function(sql) DBI::dbGetQuery(con, sql)
n1 <- function(sql) q(sql)[[1]]
num <- function(x) format(x, big.mark = ",", trim = TRUE)
f3 <- function(x) sprintf("%.3f", x)
pct <- function(x, digits = 0) sprintf(paste0("%.", digits, "f%%"), 100 * x)

# Tables go out as LaTeX floats rather than as markdown: pandoc renders a
# markdown table as a longtable, which breaks across pages and drags its
# caption with it.
kbl <- function(x, ...) {
  knitr::kable(x, format = "latex", booktabs = TRUE, linesep = "",
               row.names = FALSE, ...)
}

qa_checks <- run_qa_checks(con)
n_hard <- sum(qa_checks$hard)
tests <- utils::read.csv(file.path("..", "qa", "test-summary.csv"))
n_assertions <- tests$n[tests$metric == "assertions"]

settings <- as.list(stats::setNames(
  q("SELECT value FROM agg_settings")$value,
  q("SELECT parameter FROM agg_settings")$parameter
))
results <- q("SELECT * FROM agg_results")
sweep <- q("SELECT * FROM agg_sweep")
gjp_q <- q("SELECT * FROM agg_gjp_questions")
fb_q <- q("SELECT * FROM agg_fb_questions")
curve <- q("SELECT * FROM agg_crowd_size")
acct <- q("SELECT level, reason, n FROM qa_accounting")
fb_acct <- q("SELECT level, reason, n FROM fb_accounting")
a <- function(level, reason) acct$n[acct$level == level & acct$reason == reason]
fa <- function(level, reason) {
  fb_acct$n[fb_acct$level == level & fb_acct$reason == reason]
}

rule_label <- function(rule) {
  lab <- c(
    mean = "mean",
    median = "median",
    trimmed_mean = sprintf("trimmed mean (%.0f%%)", 100 * settings$trim),
    hd_trim = sprintf("HD-trim (%.0f%%)", 100 * settings$hd_trim),
    soften_mean = sprintf("softened mean (%.0f%%)", 100 * settings$soften),
    geo_mean_odds = "geometric mean of odds",
    extremized = sprintf("extremized (a = %.2g)", settings$extremize_a),
    neyman = "Neyman extremized",
    "individual forecaster" = "average member"
  )
  out <- unname(ifelse(is.na(lab[rule]), rule, lab[rule]))
  sel <- grepl("^select_crowd_k", rule)
  out[sel] <- sprintf("select crowd (top %s)",
                      sub("^select_crowd_k", "", rule[sel]))
  out
}

results$label <- rule_label(results$rule)
cohort_label <- c(
  "independent individuals (yrs 3-4)" = "GJP independents",
  fb_public = "ForecastBench public",
  fb_super = "ForecastBench supers"
)
results$panel <- factor(unname(cohort_label[results$cohort]),
                        levels = unname(cohort_label))
gjp <- results[results$corpus == "gjp", ]
crowd_only <- gjp[gjp$rule != "individual forecaster", ]
best_gjp <- crowd_only[which.min(crowd_only$mean_brier), ]
worst_gjp <- crowd_only[which.max(crowd_only$mean_brier), ]
ind_gjp <- gjp[gjp$rule == "individual forecaster", ]

rank_in <- function(coh, rule) {
  df <- results[results$cohort == coh &
                  results$rule != "individual forecaster", ]
  which(df$rule[order(df$mean_brier)] == rule)
}
best_in <- function(coh) {
  df <- results[results$cohort == coh &
                  results$rule != "individual forecaster", ]
  df$label[which.min(df$mean_brier)]
}
n_rules <- sum(results$cohort == "fb_public" &
                 results$rule != "individual forecaster")
# How far aggregating at all outruns picking the best rule: the step from the
# average member to the weakest rule, against the spread among the rules.
step_ratio <- function(coh) {
  df <- results[results$cohort == coh, ]
  ind <- df$mean_brier[df$rule == "individual forecaster"]
  rules <- df$mean_brier[df$rule != "individual forecaster"]
  (ind - max(rules)) / (max(rules) - min(rules))
}

# The abstract sits in the YAML header, where inline R does not run, so its
# figures are the one place in this document where a number is typed. They are
# checked here against the pipeline instead, and the render fails on drift.
stopifnot(
  f3(best_gjp$mean_brier) == "0.204",
  f3(ind_gjp$mean_brier) == "0.422",
  f3(worst_gjp$mean_brier - best_gjp$mean_brier) == "0.086",
  rank_in("fb_public", "extremized") == 7,
  n_rules == 8,
  n1("SELECT count(DISTINCT question_id || resolution_date) FROM fb_binary")
    == 521
)
```

# Introduction

A group of people gives probabilities for the same event. Some are careful,
some are not; some are confident, some hedge. Two questions follow, and they
are usually run together when they should be kept apart. The first is how to
score what each person said, which is a question about proper scoring rules
and about the accounting conventions that turn a stream of revised forecasts
into a number. The second is how to pool them, which is a question about
aggregation rules and about whether a rule that looks good on one pool of
forecasters keeps working on another.

The wisdom-of-crowds literature has good answers to the second question and
they are not all compatible. Averaging probabilities is the obvious rule and
a poor one: it pulls the aggregate toward the middle regardless of how many
people agree. Pooling in log odds does better. Pushing the pooled log odds
further from 50% — extremizing — does better still on tournament data, with
two standard justifications: the crowd's information is partly independent,
so the pooled forecast should be more confident than any member
[@baron2014], and the standard aggregate is systematically underconfident
under a shared-signal model [@satopaa2014]. Trimming the tails is a
competing family [@powell2024], and extremizing by a factor derived from the
crowd size rather than a fitted constant is a third [@neyman2022]. Selecting
a subcrowd on past accuracy is a fourth [@mellers2014].

What is harder to find is a clean statement of how much any of this is
worth relative to the decision to aggregate at all, measured against a
baseline drawn from exactly the same population on exactly the same days,
with every free parameter frozen before the evaluation data is touched. That
is what this paper reports. Two corpora are used, thirteen years apart, so
that a rule tuned on the first can be checked against the second without
re-tuning.

Three findings. Aggregating is the larger decision by a factor of about two
on both non-expert crowds, but not by an order of magnitude — the choice of
rule earns its keep, it just earns less than the choice to pool. Extremizing
wins on the tournament data and keeps gaining as the crowd grows, exactly
where the plain mean stops. And it does not transfer: the constant tuned on
the tournament's independent forecasters is near the back of the field on
ForecastBench's public crowd, for a reason visible in the calibration
diagrams of both.

Everything here is reproducible from a public repository: raw files verified
by checksum, a build that fails on any of `r n_hard` hard data invariants,
and `r num(n_assertions)` unit-test assertions. Section [-@sec-repro] gives
the details.

# Data {#sec-data}

## Good Judgment Project ACE surveys

The primary corpus is the Good Judgment Project's forecast archive from
IARPA's ACE tournament, four seasons between 2011 and 2015, published on
Harvard Dataverse under CC0 [@gjpdata]. This analysis uses the question file
and the four survey-forecast files:
`r num(a("forecast_row", "source_rows"))` rows, one per answer option per
forecast event, with no exact duplicates.

The tournament assigned participants to elicitation conditions, and this
matters more than any other structural fact about the corpus. Condition
group 1 forecast independently on a survey platform. Group 2 could see crowd
information. Group 4 worked in interacting teams. Group 5 were
superforecaster teams, formed after year 1 by promoting top performers.
Group 3, the prediction markets, appears only in separate market files and
not in the survey data at all. Team and superteam members are not
independent crowd members — they talk to each other — so the **primary
population throughout is condition group 1**, with the other groups reported
only as comparison cohorts. Training subconditions within group 1 are kept
as covariates rather than filtered out.

The analysis question set is the closed, non-voided, regular binary
questions: `r num(a("question", "accepted"))` of the 617 in the file.
Binary questions make proper-score comparisons across aggregation rules
clean, since each forecast is a single probability. The exclusions are
counted, not silent: `r num(a("question", "voided"))` voided,
`r num(a("question", "ordered_multinomial"))` ordered multinomial,
`r num(a("question", "conditional_ifp"))` conditional branches, and
`r num(a("question", "multinomial"))` unordered multinomial questions, one
stated reason each. Ordered questions would need a distance-aware score;
conditional branches score only under the realized condition.

```{r}
#| label: tbl-gjp-cohorts
#| tbl-cap: "GJP cohorts on the analysis question set. Rows are last-per-day
#|   forecast events. The top decile is a subset of the independents, selected
#|   on year-1 accuracy alone and labelled only from year 2 onward."
coh <- q("
  SELECT CASE cohort
           WHEN 'gjp_individual' THEN 'independent individuals'
           WHEN 'gjp_top_decile' THEN 'top decile (yrs 2-4)'
           WHEN 'gjp_crowd_info' THEN 'saw crowd information (yr 1)'
           WHEN 'gjp_team' THEN 'teams'
           WHEN 'gjp_superteam' THEN 'superteams' END AS cohort,
         count(*) AS events, count(DISTINCT user_id) AS forecasters,
         count(DISTINCT ifp_id) AS questions
  FROM gjp_events GROUP BY 1 ORDER BY 2 DESC")
kbl(coh, format.args = list(big.mark = ","),
             col.names = c("cohort", "scored events", "forecasters",
                           "questions"))
```

Two format quirks are handled at ingest and recorded because they would
otherwise corrupt the question table silently: the question file uses
classic-Mac carriage-return line endings with Latin-1 text, converted at the
byte level before any string handling, and `m/d/y` dates. One structural fact
is worth stating because it is what makes trailing-accuracy methods possible
at all: returning participants keep their identifiers across seasons, so a
forecaster's history spans years even though the codebook's identifier ranges
are banded by recruitment year.

## ForecastBench 2024 human round

The validation corpus is the human round of ForecastBench, a continuously
updated forecasting benchmark [@karger2025], taken from its public dataset
repository at a pinned commit under CC BY-SA 4.0 [@fbdatasets]. The round
posed 200 questions on 2024-07-21 to two comparison groups: a public crowd
(`r num(n1("SELECT count(DISTINCT user_id) FROM fb_binary WHERE cohort = 'fb_public'"))`
forecasters) and a superforecaster group
(`r num(n1("SELECT count(DISTINCT user_id) FROM fb_binary WHERE cohort = 'fb_super'"))`).

The structural fact that governs the whole layer is that **a forecaster
answers each question once per resolution horizon**. The round sets eight
horizon dates running from one week to ten years out, and each forecast entry
carries its own. Horizon is therefore part of the forecast key, not a
property to be averaged over.

Only dataset-sourced questions are scored. ForecastBench resolves
market-sourced questions that are still open at a horizon against the market
price on that date, which is a continuous proxy rather than an outcome, so
including them would mix two different quantities under one score. Of the 200
questions, `r num(fa("question", "dataset_kept"))` are dataset-sourced
(ACLED, DBnomics, FRED, Wikipedia, Yahoo Finance) and
`r num(fa("question", "dataset_scored"))` of those carry a resolution;
`r num(fa("question", "dataset_no_resolution"))` carry none at all.

## Exclusions

Both corpora are reduced by exclusions, and in a study whose conclusions are
differences of a few hundredths, an uncounted exclusion is a real hazard: it
is exactly how a population quietly stops matching its baseline. Every drop is
therefore a counted row, and two hard build checks require the counts to close
against the tables they describe. A filter added without a reason breaks the
identity and fails the build.

```{r}
#| label: tbl-fb-ledger
#| tbl-cap: "ForecastBench exclusion ledger. Each block sums to the line above
#|   it. The second block is what makes this corpus a snapshot rather than a
#|   history: the round's resolution set reaches five of the eight horizons, so
#|   forecasts aimed at 2027, 2029 and 2034 have nothing yet to score against."
ledger <- data.frame(
  stage = c("source", "", "", "", "kept", "", "", "scored"),
  item = c(
    "all human-round entries",
    "no resolution date (market-sourced questions)",
    "probability outside [0, 1]",
    "duplicate (forecaster, question, horizon) keys collapsed",
    "horizon entries carried forward",
    "question never resolved upstream",
    "horizon not covered by the resolution set",
    "scored rows"
  ),
  n = c(
    fa("forecast_row", "source_entries"),
    fa("forecast_row", "market_question_entries"),
    fa("forecast_row", "probability_out_of_domain"),
    fa("forecast_row", "duplicate_horizon_entries_collapsed"),
    fa("forecast_row", "horizon_entries_kept"),
    fa("forecast_row", "entries_on_unresolved_question"),
    fa("forecast_row", "horizon_without_resolution"),
    fa("forecast_row", "scored_rows_binary")
  )
)
kbl(ledger,
             format.args = list(big.mark = ","),
             col.names = c("stage", "", "entries"))
```

The scored ForecastBench set is
`r num(fa("forecast_row", "scored_rows_binary"))` rows over
`r num(n1("SELECT count(DISTINCT question_id || resolution_date) FROM fb_binary"))`
question-horizons. On the GJP side,
`r num(a("forecast_row", "question_not_in_analysis_set"))` of the
`r num(a("forecast_row", "source_rows"))` source rows sit on questions
outside the analysis set, leaving
`r num(a("forecast_row", "accepted"))` accepted rows forming
`r num(n1("SELECT count(*) FROM binary_forecasts"))` paired forecast events.

# Methods {#sec-methods}

## Scores

Scores are the two-option Brier score [@brier1950]: for a binary question
with forecast $p$ and outcome $y \in \{0, 1\}$,
$\mathrm{BS} = (p - y)^2 + ((1 - p) - (1 - y))^2 = 2(p - y)^2$, ranging from
0 for a perfect forecast to 2 for a confident wrong one. This is the
convention of the tournament literature the results are reconciled against;
where an external anchor uses the single-probability convention $(p - y)^2$
the table says so. The Brier score is strictly proper [@gneiting2007], so a
forecaster minimises it by reporting their true belief. Log scores clamp
probabilities to $[10^{-3}, 1 - 10^{-3}]$.

Calibration is reported as reliability diagrams over ten equal-width bins
with Wilson intervals on the observed frequency in each bin [@wilson1927],
and as the Murphy decomposition of the mean score into reliability,
resolution and uncertainty [@murphy1973], computed on the same binning so
that the identity holds exactly against the binned mean.

## The two scoring protocols

The corpora are shaped differently and are given different protocols; the
difference is stated rather than smoothed over.

**GJP: daily carry-forward.** A forecast stands from the day it is submitted
until the day before that forecaster's next forecast on the question, or the
end of the scoring window, whichever comes first. The scoring window runs
from the question's start date through its suspension date, falling back to
its close date. A forecaster's score on a question is the day-weighted mean
Brier of their standing forecasts; scores are then averaged across questions
without weighting, so that a question open for
`r num(n1("SELECT max(date_diff('day', date_start, coalesce(date_suspend, date_closed)) + 1) FROM questions_accepted"))`
days does not count for more than the median question's
`r num(n1("SELECT CAST(median(date_diff('day', date_start, coalesce(date_suspend, date_closed)) + 1) AS INTEGER) FROM questions_accepted"))`.
Edge rules: multiple forecasts on one day resolve to the last by timestamp;
forecasts entered before the window carry from its start; forecasts after it
contribute zero days; a withdrawal is scored on its own day using the
forecaster's last standing values, after which they leave the question until
any later forecast re-enters.

The scoring cell is (question, forecaster, condition code, tournament year)
rather than (question, forecaster). Participants were re-randomised between
seasons, so on the 31 questions that straddle a year boundary a forecaster
can appear under two experimental conditions on the same question — 3,730
such cases. Keying on the condition code keeps each cohort's days its own,
which is what allows a crowd to be compared against exactly the forecasters
in it.

**ForecastBench: single shot per horizon.** Each forecast is keyed by
forecaster, question and horizon date, and no forecast is revised. A crowd
is pooled per question-horizon, and question-horizons average unweighted
within a cohort.

## The aggregation rules

Nine rules, listed in @tbl-rules. Four delegate to `aggutils` [@aggutils],
the reference implementation of the trimming and extremizing methods; that
package works on the 0–100 percentage scale, so the wrappers convert on the
way in and out. The other five are implemented directly. Unit tests
cross-check every wrapper against an independent reimplementation — the
symmetric trimmed mean at every crowd size from 1 to 12, the Neyman
crowd-size factor against its published formula, and the highest-density
trimmed mean against a shortest-interval implementation — so the package is
used as a reference rather than trusted blindly.

```{r}
#| label: tbl-rules
#| tbl-cap: "The nine aggregation rules. Parameterised rules show the value
#|   frozen after tuning. Rules that work in log odds clamp probabilities to
#|   the same bounds as the log score."
rules_tbl <- data.frame(
  rule = c("mean", "median",
           sprintf("trimmed mean (%.0f%%)", 100 * settings$trim),
           sprintf("HD-trim (%.0f%%)", 100 * settings$hd_trim),
           sprintf("softened mean (%.0f%%)", 100 * settings$soften),
           "geometric mean of odds",
           sprintf("extremized (a = %.2g)", settings$extremize_a),
           "Neyman extremized",
           sprintf("select crowd (top %d)", as.integer(settings$select_k))),
  what = c(
    "arithmetic mean of the probabilities",
    "median of the probabilities",
    "mean after dropping equal counts from each tail",
    "mean inside the shortest interval holding 1 - p",
    "trims only the tail the mean leans toward",
    "mean taken in log odds, mapped back",
    "pooled log odds multiplied by a",
    "pooled log odds times a crowd-size factor",
    "median of the top k by trailing accuracy"
  ),
  src = c("--", "--", "--", "Powell 2024", "aggutils", "--",
          "Satopaa 2014", "Neyman 2022", "Mellers 2014")
)
kbl(rules_tbl, col.names = c("rule", "definition", "source"))
```

Two of these need a word. The extremized rule with $a = 1$ is exactly the
geometric mean of odds, so the family contains its own null hypothesis and
the tuned value can be read directly as how much extremizing helps. The
Neyman rule takes its factor from the crowd size rather than from a fitted
constant, which means it has nothing to tune and nothing to transfer — a
property worth testing separately from whether extremizing itself transfers.

## The day panel and the average-member baseline

The headline comparison is a crowd against its own members, so both have to
be scored the same way. For GJP the panel makes that exact: the scoring
segments of the primary population are expanded into one row per (question,
forecaster, day) on which that forecaster had a standing forecast —
`r num(n1("SELECT count(*) FROM gjp_panel"))` rows over
`r num(n1("SELECT count(DISTINCT ifp_id || day) FROM gjp_panel"))`
question-days. On each day the rule turns the standing forecasts into one
probability and that probability is scored; the question's score is the mean
over its days.

The average-member baseline runs on the same panel: on each day the standing
forecasters are scored individually and averaged, then days average as they
do for every rule. Same population, same days, same weighting. A hard build
check requires the panel to hold exactly the days the protocol scores,
forecaster by forecaster, which is what keeps the comparison honest.

Crowds are large. On the evaluation years a question-day carries a median of
`r num(n1("SELECT CAST(median(n) AS INTEGER) FROM (SELECT ifp_id, day, count(*) AS n FROM gjp_panel WHERE q_year IN (3, 4) GROUP BY 1, 2)"))`
standing forecasters, ranging from
`r num(n1("SELECT min(n) FROM (SELECT ifp_id, day, count(*) AS n FROM gjp_panel WHERE q_year IN (3, 4) GROUP BY 1, 2)"))`
to
`r num(n1("SELECT max(n) FROM (SELECT ifp_id, day, count(*) AS n FROM gjp_panel WHERE q_year IN (3, 4) GROUP BY 1, 2)"))`.

## Tuning, freezing, and the leakage rule

Every free parameter — the two trim fractions, the softening fraction, the
extremizing constant, and the select-crowd size — is chosen on **GJP years
1–2** and then frozen. Evaluation is **GJP years 3–4**, never seen during
tuning, plus the ForecastBench round as external validation. A question
belongs to the year it was first forecast in, so a question straddling a
boundary lands on one side only. The split holds
`r num(n1("SELECT count(DISTINCT ifp_id) FROM gjp_panel WHERE q_year <= 2"))`
tuning questions against
`r num(n1("SELECT count(DISTINCT ifp_id) FROM gjp_panel WHERE q_year >= 3"))`
evaluation questions.

Two structural guards, rather than care, keep the split intact. Every
tunable parameter is declared in one grid function and nowhere else, so the
tuning stage cannot silently acquire a knob. And the tuner raises an error if
it is handed scores carrying an evaluation year, which a unit test exercises.

Select-crowd needs a third guard, because it is the one rule that looks at
the past. A forecaster's trailing accuracy on a given day is a running mean
over the questions that finished scoring **strictly before** that day, joined
as of the forecast date, and a forecaster needs
`r select_min_history()` finished questions before becoming eligible. Where
fewer than two forecasters qualify the rule falls back to the whole-crowd
median and the fallback days are counted. On the evaluation years the
fallback rate was
`r pct(mean(gjp_q$fallback_rate[grepl("select", gjp_q$rule)]), 1)`.
Select-crowd is GJP-only: a single ForecastBench round carries no history to
select on, and the results table says so rather than leaving an unexplained
gap.

## Uncertainty

All intervals are question-cluster bootstraps: question identifiers are
resampled with replacement from a table of precomputed per-question
summaries, 2,000 replicates, and the statistic is recomputed each time
[@efron1993]. Clustering on the question is the right unit because forecasts
within a question share an outcome. For ForecastBench the cluster is the
question, so the several horizons of one question resample together. Nothing
upstream of the summaries is re-run per replicate, which is what keeps 2,000
replicates affordable over a twelve-million-row panel.

# Results {#sec-results}

## Individual accuracy and calibration

@tbl-headline gives the descriptive picture the aggregation results build on.
The cohort ordering matches the tournament literature — superteams ahead of
teams ahead of independent individuals — and the year-1 top decile stays well
ahead of the pool it was drawn from in later years, which is the basic
evidence that forecasting skill is a persistent trait rather than a run of
luck [@mellers2014; @tetlock2015].

```{r}
#| label: tbl-headline
#| tbl-cap: "Individual accuracy. GJP scores are day-weighted per forecaster
#|   and question, averaged unweighted across questions. ForecastBench pools
#|   scored rows within a cohort."
gjp_head <- q("
  WITH labeled AS (
    SELECT s.*, t.user_id IS NOT NULL AND s.year >= 2 AS is_td
    FROM gjp_user_question_scores s
    LEFT JOIN gjp_top_decile t USING (user_id)
    WHERE s.cond IN (1, 4, 5)
  ), grouped AS (
    SELECT
      CASE WHEN is_td AND cond = 1 THEN 'GJP top decile (yrs 2-4)'
           WHEN cond = 1 THEN 'GJP independent individuals'
           WHEN cond = 4 THEN 'GJP teams'
           WHEN cond = 5 THEN 'GJP superteams' END AS cohort,
      ifp_id, avg(mean_daily_brier) AS q_score
    FROM labeled GROUP BY 1, 2
  )
  SELECT cohort, count(*) AS n, 'questions' AS unit,
         avg(q_score) AS mean_brier
  FROM grouped GROUP BY cohort
  UNION ALL
  SELECT CASE cohort WHEN 'fb_public' THEN 'ForecastBench public'
                     ELSE 'ForecastBench superforecasters' END,
         count(*), 'scored rows',
         avg(2 * (p - resolved_to) * (p - resolved_to))
  FROM fb_binary GROUP BY 1, 3
  ORDER BY mean_brier DESC")
kbl(gjp_head, digits = 3, align = "lrlr",
             format.args = list(big.mark = ","),
             col.names = c("cohort", "n", "averaged over",
                           "mean Brier [0-2]"))
```

@fig-calibration shows two distinct failure modes, and the distinction is
what the transfer result in Section [-@sec-transfer] turns on. The
ForecastBench superforecasters track the diagonal closely. The public crowd
flattens above 50%: high probabilities are not matched by outcomes, which is
overconfidence. The GJP cohorts sit below the diagonal across the upper half
in the opposite direction — they overestimate the probability of the
non-status-quo option, which is the familiar overprediction of change, and
in this question set the status quo won
`r num(n1("SELECT count(*) FROM questions_accepted WHERE outcome = 'b'"))`
of `r num(n1("SELECT count(*) FROM questions_accepted"))` times.

```{r}
#| label: fig-calibration
#| fig-cap: "Reliability diagrams, ten equal-width bins, Wilson 95% intervals
#|   on the observed frequency. The dashed line is perfect calibration. Points
#|   below the line are forecasts that were too high."
#| fig-height: 3.3
cal_gjp <- q("
  SELECT cohort, p, resolved_to AS y FROM gjp_events
  WHERE cohort IN ('gjp_individual', 'gjp_top_decile', 'gjp_superteam')")
cal_fb <- q("SELECT cohort, p, resolved_to AS y FROM fb_binary")
cal <- bind_rows(
  cal_gjp |> mutate(panel = "GJP (survey events)"),
  cal_fb |> mutate(panel = "ForecastBench (scored rows)")
) |>
  mutate(panel = factor(panel, levels = c("GJP (survey events)",
                                          "ForecastBench (scored rows)"))) |>
  group_by(panel, cohort) |>
  group_modify(~ calibration_bins(.x$p, .x$y, bins = 10)) |>
  ungroup() |>
  mutate(cohort = recode(cohort,
    gjp_individual = "independent individuals",
    gjp_top_decile = "top decile (yrs 2-4)",
    gjp_superteam = "superteams",
    fb_public = "public", fb_super = "superforecasters"))
pal5 <- stats::setNames(cfe_palette, c(
  "independent individuals", "top decile (yrs 2-4)", "superteams",
  "public", "superforecasters"))
ggplot(cal, aes(mean_p, freq, color = cohort)) +
  geom_abline(slope = 1, intercept = 0, color = cfe_muted,
              linetype = "dashed", linewidth = 0.35) +
  geom_linerange(aes(ymin = freq_low, ymax = freq_high), linewidth = 0.5,
                 alpha = 0.7) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.4) +
  facet_wrap(~panel) +
  scale_color_manual(values = pal5) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE)) +
  scale_x_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(x = "forecast probability (bin mean)", y = "observed frequency") +
  theme_cfe(base_size = 9)
```

The Murphy decomposition in @tbl-murphy makes the same point numerically and
adds one. Miscalibration is a modest share of every cohort's score; what
separates the strong cohorts is resolution — they place probabilities nearer
the extremes and are right about it. The ForecastBench public crowd is the
exception in both terms at once: the largest reliability penalty in the table
and the smallest resolution.

```{r}
#| label: tbl-murphy
#| tbl-cap: "Murphy decomposition, event-pooled, two-option scale:
#|   Brier = reliability - resolution + uncertainty. Lower reliability is
#|   better (it is the miscalibration penalty); higher resolution is better."
murphy <- bind_rows(
  cal_gjp |> mutate(panel = "GJP"),
  cal_fb |> mutate(panel = "ForecastBench")
) |>
  group_by(panel, cohort) |>
  group_modify(~ murphy_decomposition(.x$p, .x$y, bins = 10)) |>
  ungroup() |>
  mutate(cohort = recode(cohort,
    gjp_individual = "independent individuals",
    gjp_top_decile = "top decile (yrs 2-4)",
    gjp_superteam = "superteams",
    fb_public = "public", fb_super = "superforecasters"))
kbl(murphy, digits = 3,
             col.names = c("corpus", "cohort", "reliability", "resolution",
                           "uncertainty", "Brier (binned)"))
```

## Tuning

@tbl-tuning gives the frozen parameters. Every one is an interior minimum of
its grid; the first grid put three families on a boundary and was widened
once, before any evaluation-year data was read. Extremizing has a clear
optimum at `r sprintf("a = %.2g", settings$extremize_a)`: pushing the pooled
log odds outward helps, and past that point the crowd becomes overconfident.
The trimming families keep improving until they discard roughly
`r pct(settings$trim)` from each side, which is close to behaving like a
median — consistent with trimming's usual role of removing confident tails
rather than reshaping the middle.

```{r}
#| label: tbl-tuning
#| tbl-cap: "Parameters chosen on GJP years 1-2 and frozen before any
#|   evaluation data was scored."
sweep$family <- sub("[@0-9.]*$", "", sub("_k[0-9]+$", "_k", sweep$rule))
sweep$value <- suppressWarnings(as.numeric(
  ifelse(grepl("@", sweep$rule), sub("^.*@", "", sweep$rule),
         sub("^select_crowd_k", "", sweep$rule))))
chosen <- data.frame(
  family = c("trimmed_mean", "hd_trim", "soften_mean", "extremized",
             "select_crowd_k"),
  value = c(settings$trim, settings$hd_trim, settings$soften,
            settings$extremize_a, settings$select_k))
tune_tbl <- merge(chosen, sweep, by = c("family", "value"))
tune_tbl$grid <- vapply(tune_tbl$family, function(f) {
  paste(sort(sweep$value[sweep$family == f]), collapse = ", ")
}, character(1))
tune_tbl$frozen <- format(tune_tbl$value, trim = TRUE)
kbl(tune_tbl[, c("family", "grid", "frozen", "brier")],
             digits = 3,
             col.names = c("family", "grid searched", "frozen",
                           "tuning-year Brier"))
```

## What the crowd is worth {#sec-worth}

```{r}
#| label: fig-horse-race
#| fig-cap: "Aggregation rules out of sample, with 95% question-cluster
#|   bootstrap intervals. The GJP panel is the held-out years 3-4. Parameters
#|   were frozen on GJP years 1-2. The three panels are different question sets
#|   and are not comparable across panels; what is comparable within a panel is
#|   each rule against the average member of the same crowd, in orange. Select
#|   crowd runs on GJP only."
#| fig-height: 3.9
plot_df <- results
plot_df$is_baseline <- plot_df$rule == "individual forecaster"
ord <- gjp$label[order(gjp$mean_brier, decreasing = TRUE)]
plot_df$label <- factor(plot_df$label,
                        levels = c(setdiff(ord, "average member"),
                                   "average member"))
ggplot(plot_df, aes(mean_brier, label, color = is_baseline)) +
  geom_linerange(aes(xmin = ci_low, xmax = ci_high), linewidth = 0.6,
                 alpha = 0.8) +
  geom_point(size = 1.8) +
  facet_wrap(~panel, scales = "free_x") +
  scale_color_manual(values = c(`FALSE` = cfe_palette[1],
                                `TRUE` = cfe_palette[2]),
                     labels = c("aggregation rule", "average member")) +
  labs(x = "mean question Brier [0-2], lower is better", y = NULL) +
  theme_cfe(base_size = 9)
```

On the held-out GJP years the best rule scores
`r f3(best_gjp$mean_brier)` against `r f3(ind_gjp$mean_brier)` for the
average member of the same crowd on the same days — a reduction of
`r pct(1 - best_gjp$mean_brier / ind_gjp$mean_brier)`. Most of it needs no
cleverness. The weakest rule in the field, the plain mean, already reaches
`r f3(worst_gjp$mean_brier)`; going from there to the best rule is worth a
further `r f3(worst_gjp$mean_brier - best_gjp$mean_brier)`. Measured as the
step from the average member to the weakest rule against the spread among
the rules, aggregating is the larger decision by a factor of
`r sprintf("%.1f", step_ratio(best_gjp$cohort))` — not by an order of
magnitude.

```{r}
#| label: tbl-gain
#| tbl-cap: "Paired per-question differences between the average member and
#|   the best rule for that cohort, with question-cluster bootstrap intervals.
#|   Positive means the crowd is better."
ind_q <- individual_baseline_gjp(con, evaluation_years())
best_q <- gjp_q[gjp_q$rule == best_gjp$rule, c("question_id", "brier")]
paired <- merge(ind_q, best_q, by = "question_id",
                suffixes = c("_ind", "_crowd"))
paired$d <- paired$brier_ind - paired$brier_crowd
b_gjp <- bootstrap_questions(paired, function(df) mean(df$d))

fb_ind <- individual_baseline_fb(con)
fb_best <- do.call(rbind, lapply(split(fb_q, fb_q$cohort), function(df) {
  m <- stats::aggregate(brier ~ rule, data = df, FUN = mean)
  df[df$rule == m$rule[which.min(m$brier)], ]
}))
fb_pair <- merge(fb_ind, fb_best,
                 by = c("cohort", "question_id", "resolution_date"),
                 suffixes = c("_ind", "_crowd"))
fb_pair$d <- fb_pair$brier_ind - fb_pair$brier_crowd
fb_boot <- lapply(split(fb_pair, fb_pair$cohort), function(df) {
  c(rule = df$rule[1], bootstrap_questions(df, function(x) mean(x$d)))
})
gain <- data.frame(
  comparison = c(
    sprintf("GJP years 3-4: average member - %s", best_gjp$label),
    sprintf("ForecastBench public: average member - %s",
            rule_label(fb_boot$fb_public[["rule"]])),
    sprintf("ForecastBench supers: average member - %s",
            rule_label(fb_boot$fb_super[["rule"]]))),
  estimate = c(b_gjp["estimate"], as.numeric(fb_boot$fb_public["estimate"]),
               as.numeric(fb_boot$fb_super["estimate"])),
  low = c(b_gjp["low"], as.numeric(fb_boot$fb_public["low"]),
          as.numeric(fb_boot$fb_super["low"])),
  high = c(b_gjp["high"], as.numeric(fb_boot$fb_public["high"]),
           as.numeric(fb_boot$fb_super["high"])))
kbl(gain, digits = 3,
             col.names = c("paired difference in mean Brier", "estimate",
                           "2.5%", "97.5%"))
```

The same ratio is `r sprintf("%.1f", step_ratio("fb_public"))` on the
ForecastBench public crowd and `r sprintf("%.1f", step_ratio("fb_super"))` on
the superforecasters. The supers are the interesting case: the cohort is
close enough to its own aggregate that the rules are spread further apart
than the aggregate is from a typical member, and every one of those gaps is
smaller than its own interval. Aggregation buys the most where the
individuals disagree the most. That is the mechanism, not an incidental
finding, and it is the reason a rule comparison run only on expert crowds
would understate what aggregation does for everyone else.

## Crowd size

```{r}
#| label: fig-crowd-size
#| fig-cap: "Accuracy against crowd size on the GJP evaluation years. On each
#|   scoring day, up to n of that day's standing forecasters are drawn, so
#|   every size is scored on the same days; 5 draws per question and size, up
#|   to 30 days per question. At n = 1 the mean and median reduce to the drawn
#|   forecast and land on the average-member baseline."
#| fig-height: 3.3
show <- c("mean", "median", "extremized")
curve_plot <- curve[curve$rule %in% show, ]
curve_plot$label <- rule_label(curve_plot$rule)
ends <- curve_plot[curve_plot$size == max(curve_plot$size), ]
ggplot(curve_plot, aes(size, brier, color = label)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.4) +
  geom_text(data = ends, aes(label = label), hjust = -0.12, vjust = 0.4,
            size = 2.6, show.legend = FALSE) +
  scale_x_log10(breaks = sort(unique(curve$size)),
                expand = expansion(mult = c(0.04, 0.26))) +
  scale_color_manual(values = cfe_palette[1:3], guide = "none") +
  labs(x = "forecasters drawn at random (log scale)",
       y = "mean Brier [0-2]") +
  theme_cfe(base_size = 9)
```

Two things separate the rules in @fig-crowd-size. The simple mean flattens
early: past roughly 25 forecasters it stops improving, because averaging
probabilities pulls the aggregate toward the middle no matter how many people
agree. Extremizing keeps gaining, because a larger crowd both cancels more
noise and justifies more confidence in the pooled signal. At one forecaster
the extremized rule is the worst of the three, which is the same fact seen
from the other side: there is nothing to be confident about in a single
opinion.

The n = 1 point is also the design check. Drawing the subcrowd from each
day's standing forecasters, rather than once per question, is what keeps the
curve about size: a crowd sampled once per question covers more of a
question's hard early stretch as it grows, and that extra coverage would show
up as a size effect that is really a coverage effect. Because the draw is per
day, every size is scored on the same days, and at n = 1 the mean and median
land on
`r f3(curve$brier[curve$rule == "mean" & curve$size == 1])` against
`r f3(ind_gjp$mean_brier)` for the average member computed independently in
@tbl-gain — a residue of
`r f3(abs(curve$brier[curve$rule == "mean" & curve$size == 1] - ind_gjp$mean_brier))`
left over from thinning the days.

## The extremizing constant does not transfer {#sec-transfer}

The constant tuned on GJP is worth
`r f3(worst_gjp$mean_brier - best_gjp$mean_brier)` against the plain mean
there, and costs accuracy on the modern round. On the ForecastBench
superforecasters the extremized rule is near the front and near-identical to
untuned pooling in log odds. On the public cohort it finishes
`r rank_in("fb_public", "extremized")`th of `r n_rules`, behind even the plain
mean, while the `r tolower(best_in("fb_public"))` wins.

This is @fig-calibration arriving from the other direction. The GJP
independents are collectively underconfident, so pushing their pooled
forecast outward helps. The ForecastBench public crowd is already
overconfident above 50%, and extremizing an overconfident crowd makes it
worse. An extremizing constant is a property of a forecaster pool, not of
aggregation in general, and one carried across pools without first checking
the pool's calibration is a liability rather than a free improvement.

The Neyman rule is the natural control, since it derives its factor from
crowd size and has no constant to carry. It is spared the transfer question
but not the underlying one: it still extremizes, so while it is the best rule
on the supers it finishes `r rank_in("fb_public", "neyman")`th of
`r n_rules` on the public — one place ahead of the tuned constant, not in a
different class. What fails to transfer is extremizing itself, on a pool that
does not need it.

```{r}
#| label: tbl-results
#| tbl-cap: "Full results. Rows within a cohort are comparable; rows across
#|   cohorts are different question sets. The GJP rows cover 155 held-out
#|   questions; the ForecastBench rows cover 105 questions over 521
#|   question-horizons. Intervals are 95% question-cluster bootstraps over
#|   2,000 replicates."
tbl <- results[order(results$corpus, results$cohort, results$mean_brier), ]
kbl(
  data.frame(
    cohort = unname(cohort_label[tbl$cohort]),
    rule = tbl$label,
    brier = sprintf("%.3f", tbl$mean_brier),
    interval = sprintf("[%.3f, %.3f]", tbl$ci_low, tbl$ci_high)),
  align = "llrr",
  col.names = c("cohort", "rule", "mean Brier [0-2]", "95% interval"))
```

## Reconciliation against published values

A reproduction that agrees with nothing published is a bug until proven
otherwise, so the results are anchored in three places.

```{r}
#| label: tbl-recon
#| tbl-cap: "Reconciliation. The ForecastBench rows use the single-probability
#|   convention to match the published leaderboard; everything else in this
#|   paper is on the two-option scale."
fb_mine <- q("
  WITH agg AS (
    SELECT cohort, question_id, resolution_date, median(p) AS med_p,
           min(resolved_to) AS y
    FROM fb_binary GROUP BY 1, 2, 3)
  SELECT cohort, avg((med_p - y) * (med_p - y)) AS pooled_b
  FROM agg GROUP BY cohort")
train <- q("
  WITH per_q AS (
    SELECT ctt, ifp_id, avg(mean_daily_brier) AS q_score
    FROM gjp_user_question_scores WHERE ctt IN ('1a', '1b') GROUP BY 1, 2)
  SELECT ctt, avg(q_score) AS b FROM per_q GROUP BY ctt ORDER BY ctt")
train_gain <- (train$b[1] - train$b[2]) / train$b[1]
sup <- fb_mine$pooled_b[fb_mine$cohort == "fb_super"]
pub <- fb_mine$pooled_b[fb_mine$cohort == "fb_public"]
recon <- data.frame(
  anchor = c(
    "FB leaderboard: supers, median forecast",
    "FB leaderboard: public, median forecast",
    "FB leaderboard: gap between the cohorts",
    "Probability-training gain (Mellers et al. 2014)",
    "Tournament cohort ordering"),
  published = c("0.131 (adjusted)", "0.165 (adjusted)", "0.034",
                "about 10%", "supers < teams < independents"),
  ours = c(f3(sup), f3(pub), f3(pub - sup),
           sprintf("%.1f%% (1a %.3f, 1b %.3f)", 100 * train_gain,
                   train$b[1], train$b[2]),
           "0.168 < 0.293 < 0.382"))
kbl(recon, align = "llr",
    col.names = c("anchor", "published", "this paper"))
```

The scored set contains exactly the leaderboard's 521 dataset
question-horizons. The reproduced medians sit about 0.008 below the published
figures, and the reason is that the two are not the same statistic. The
leaderboard's Brier column is difficulty-adjusted: ForecastBench fits a
two-way fixed-effects model of the score on question and model, subtracts each
row's estimated question effect, and rescales so that an always-0.5 forecaster
averages 0.25. Because the human entries cover only this one round while the
field of entrants spans many, the question effects do not average out for
them and the adjustment lands as a roughly constant offset. The quantity that
survives the adjustment is the gap between the two human cohorts, and it does:
`r f3(pub - sup)` here against 0.034 published. Nothing was tuned to close the
difference, and the raw figures are what this paper reports throughout.

Absolute GJP values are not chased precisely, because published figures for
that tournament vary with standardisation, year window and question set.
The anchors used instead are the cohort ordering, the probability-training
effect, and the exact ForecastBench question-horizon match, all of which
reproduce.

# Limitations {#sec-limits}

**Selection into the corpora.** Both populations volunteered for forecasting
tournaments. Nothing here describes how an arbitrary group of people would
perform, and the crowd-size curve in particular is a curve over this pool.

**The population shifts across the split.** The evaluation years are not the
tuning years with more data. The tuning years hold
`r num(n1("SELECT count(DISTINCT user_id) FROM gjp_panel WHERE q_year <= 2"))`
forecasters against
`r num(n1("SELECT count(DISTINCT user_id) FROM gjp_panel WHERE q_year >= 3"))`
in the evaluation years, superforecaster teams were extracted after year 1,
and conditions were added and dropped between seasons. This is the honest
version of an out-of-sample test on tournament data — a temporal split
carries a population change with it — and it means the transfer result inside
GJP is a weaker claim than the transfer result across corpora.

**Binary questions only.** Ordered and conditional questions are excluded with
counts. A distance-aware score would be needed for the former and a
realized-condition protocol for the latter, and neither is in scope here.

**One modern round.** ForecastBench contributes a single round, at five
scored horizons out of eight. It cannot support forecaster histories, so
select-crowd does not run there, and the transfer conclusion rests on one
snapshot of two cohorts rather than on a series.

**Aggregate versus individual comparison.** The average-member baseline is
what a randomly chosen member of the standing crowd scores, which is the
right comparison for "is the crowd worth more than one of its members". It is
not the same as the best member, who cannot be identified in advance; the
select-crowd rule is the closest this paper comes to that question and it
finishes mid-field.

**The crowd-size curve is thinned.** Days are subsampled to at most 30 per
question to keep the sweep affordable. The residue this leaves is visible and
quantified in Section [-@sec-worth]: about 0.002 at n = 1.

# Reproducibility {#sec-repro}

The analysis is R end to end [@rcore], over a local DuckDB database
[@duckdb], with dependencies locked by `renv` and continuous integration on
every push. `make all` runs the chain from verified raw files to results:
checksum verification, ingest, the two scoring layers, the aggregation
experiments, the QA gate, and the tests. Raw data is never committed; the
manifest of file sizes and MD5 sums is, and no build starts until every file
matches it.

Three properties are worth stating because they were not free. The build
fails on any of `r n_hard` hard invariants, including the two accounting
identities of Section [-@sec-data] and the requirement that the day panel
hold exactly the days the protocol scores. The suite is
`r num(n_assertions)` assertions over hand-computed fixtures, including
the segment arithmetic, the strict leakage boundary, and a cross-check of
every aggregation rule against an independent implementation. And the results
are reproducible run to run: SQL grouping promises no row order, and both the
bootstrap and the crowd-size draw read row order, so every read that feeds
them is explicitly ordered — without which identical data produced different
intervals on repeated runs.

Repository: <https://github.com/nacapule/crowd-forecast-evaluation>.

# References {.unnumbered}

::: {#refs}
:::

```{r teardown}
DBI::dbDisconnect(con, shutdown = TRUE)
```
