---
title: "Scoring and calibration of crowd forecasts"
subtitle: "GJP ACE surveys (2011–2015) and the ForecastBench 2024 human round"
format:
  html:
    embed-resources: true
    toc: true
    fig-width: 8
    fig-height: 4.5
execute:
  echo: false
  warning: false
---

```{r setup}
source(file.path("..", "R", "ingest.R"))
source(file.path("..", "R", "scoring.R"))
suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
con <- DBI::dbConnect(
  duckdb::duckdb(),
  dbdir = file.path("..", "data", "db", "gjp.duckdb"), read_only = TRUE
)
q <- function(sql) DBI::dbGetQuery(con, sql)

fig_dir <- "figures"
dir.create(fig_dir, showWarnings = FALSE)

pal <- c("#2a78d6", "#eb6834", "#1baf7a")
ink <- "#0b0b0b"
ink2 <- "#52514e"
muted <- "#898781"
grid_c <- "#e1e0d9"
surface <- "#fcfcfb"

theme_cfe <- function() {
  theme_minimal(base_size = 12) +
    theme(
      plot.background = element_rect(fill = surface, color = NA),
      panel.background = element_rect(fill = surface, color = NA),
      panel.grid.major = element_line(color = grid_c, linewidth = 0.3),
      panel.grid.minor = element_blank(),
      text = element_text(color = ink),
      axis.text = element_text(color = muted, size = 9),
      axis.title = element_text(color = ink2, size = 10),
      strip.text = element_text(color = ink, face = "bold", size = 10),
      legend.position = "bottom",
      legend.text = element_text(color = ink2, size = 9),
      legend.title = element_blank(),
      plot.title = element_text(face = "bold", size = 13),
      plot.subtitle = element_text(color = ink2, size = 10)
    )
}
save_fig <- function(p, name, w = 8, h = 4.5) {
  ggsave(file.path(fig_dir, name), p, width = w, height = h, bg = surface,
         dpi = 200)
  p
}
```

## Scope

This note reports the descriptive core of the project: how accurate and how
calibrated the forecasters in two public corpora are, under a precisely stated
scoring protocol. Aggregation experiments build on these tables in the next
stage.

Scores use the two-option Brier score (Brier 1950): for a binary question with
forecast $p$ and outcome $y \in \{0,1\}$, $\mathrm{BS} = 2(p-y)^2$, ranging
from 0 (perfect) to 2 (confidently wrong). This is the convention of the Good
Judgment Project literature; where an external anchor uses the
single-probability convention $(p-y)^2$ the table says so. Log scores clamp
probabilities to $[10^{-3}, 1-10^{-3}]$.

## Scoring protocols

**GJP (survey panel).** A forecast stands from its submission day until the
day before the forecaster's next forecast on that question, or the end of the
scoring window (`date_start` through `date_suspend`, falling back to
`date_closed`). Multiple forecasts on one day: the last stands. Forecasts
before the window carry from its start; forecasts after it score zero days. A
withdrawal is scored on its own day and the forecaster then exits the
question. A forecaster's question score is the day-weighted mean Brier of
their standing forecasts; cross-question averages are unweighted, so
long-lived questions do not dominate. The primary population is condition
group 1 (independent survey individuals); teams, superteams, and top-decile
forecasters (defined on year-1 accuracy only, evaluated on years 2–4) appear
as comparison cohorts.

**ForecastBench (single shot per horizon).** Humans forecast each question
once; the benchmark resolves the same forecast at several later horizon
dates. Market-sourced questions unresolved at a horizon are scored upstream
against the market price — a proxy, not an outcome — so this analysis keeps
only dataset-source questions (ACLED, DBnomics, FRED, Wikipedia, Yahoo
Finance) with genuine 0/1 resolutions: 110 of the round's 200 questions, 521
question-horizons.

## Corpus and cohorts

```{r cohort-table}
cohorts <- q("
  SELECT cohort,
         count(*) AS events,
         count(DISTINCT user_id) AS forecasters,
         count(DISTINCT ifp_id) AS questions
  FROM gjp_events GROUP BY cohort
  UNION ALL
  SELECT cohort, count(*), count(DISTINCT user_id),
         count(DISTINCT question_id)
  FROM fb_binary GROUP BY cohort
  ORDER BY cohort")
knitr::kable(
  cohorts,
  col.names = c("cohort", "scored rows", "forecasters", "questions"),
  format.args = list(big.mark = ",")
)
```

GJP rows are last-per-day forecast events on the 303 closed binary questions;
ForecastBench rows are forecaster × question × horizon scores. The full
raw-to-accepted accounting is in `qa/qa_report.md`.

## Headline accuracy

GJP: each question's cohort score is the mean over its forecasters'
day-weighted scores; questions then average unweighted. ForecastBench: the
individual column pools scored rows; the aggregate column scores the
median forecast per question-horizon (the benchmark's human-baseline
construction).

```{r headline}
gjp_head <- q("
  WITH labeled AS (
    SELECT s.*, t.user_id IS NOT NULL AND s.year >= 2 AS is_td
    FROM gjp_user_question_scores s
    LEFT JOIN gjp_top_decile t USING (user_id)
    WHERE s.cond IN (1, 4, 5)
  ), grouped AS (
    SELECT
      CASE WHEN is_td AND cond = 1 THEN 'GJP top decile (yrs 2-4)'
           WHEN cond = 1 THEN 'GJP independent individuals'
           WHEN cond = 4 THEN 'GJP teams'
           WHEN cond = 5 THEN 'GJP superteams' END AS cohort,
      ifp_id, avg(mean_daily_brier) AS q_score
    FROM labeled GROUP BY 1, 2
  )
  SELECT cohort, count(*) AS n_questions,
         avg(q_score) AS mean_brier
  FROM grouped GROUP BY cohort ORDER BY mean_brier DESC")

fb_head <- q("
  WITH agg AS (
    SELECT cohort, question_id, resolution_date,
           median(p) AS med_p, min(resolved_to) AS y
    FROM fb_binary GROUP BY 1, 2, 3
  ), ind AS (
    SELECT cohort, avg(2 * (p - resolved_to) * (p - resolved_to)) AS ind_b
    FROM fb_binary GROUP BY cohort
  )
  SELECT a.cohort, count(*) AS n_qh,
         avg(2 * (med_p - y) * (med_p - y)) AS agg_brier,
         min(i.ind_b) AS ind_brier
  FROM agg a JOIN ind i USING (cohort)
  GROUP BY a.cohort ORDER BY agg_brier DESC")

knitr::kable(
  gjp_head, digits = 3,
  col.names = c("cohort", "questions", "mean Brier [0-2]")
)
knitr::kable(
  fb_head, digits = 3,
  col.names = c("cohort", "question-horizons",
                "median-forecast Brier [0-2]", "individual Brier [0-2]")
)
```

The orderings match the tournament literature: superteams beat teams, teams
beat independent individuals, and the year-1 top decile stays well ahead of
the individual pool in later years; ForecastBench superforecasters beat the
public at both the individual and the aggregate level.

## Calibration

Reliability diagrams over forecast events (each submission once), ten
equal-width bins, Wilson 95% intervals on outcome frequency per bin.

```{r calibration, fig.height=4.2}
cal_gjp <- q("
  SELECT cohort, p, resolved_to AS y FROM gjp_events
  WHERE cohort IN ('gjp_individual', 'gjp_top_decile', 'gjp_superteam')")
cal_fb <- q("SELECT cohort, p, resolved_to AS y FROM fb_binary")

cal <- bind_rows(
  cal_gjp |> mutate(panel = "GJP (survey events)"),
  cal_fb |> mutate(panel = "ForecastBench (scored rows)")
) |>
  mutate(panel = factor(panel, levels = c(
    "GJP (survey events)", "ForecastBench (scored rows)"
  ))) |>
  group_by(panel, cohort) |>
  group_modify(~ calibration_bins(.x$p, .x$y, bins = 10)) |>
  ungroup() |>
  mutate(cohort = recode(cohort,
    gjp_individual = "independent individuals",
    gjp_top_decile = "top decile (yrs 2-4)",
    gjp_superteam = "superteams",
    fb_public = "public", fb_super = "superforecasters"
  ))

pal5 <- c(
  "independent individuals" = "#2a78d6",
  "top decile (yrs 2-4)" = "#eb6834",
  "superteams" = "#1baf7a",
  "public" = "#eda100",
  "superforecasters" = "#e87ba4"
)
p_cal <- ggplot(cal, aes(mean_p, freq, color = cohort)) +
  geom_abline(slope = 1, intercept = 0, color = muted,
              linetype = "dashed", linewidth = 0.35) +
  geom_linerange(aes(ymin = freq_low, ymax = freq_high),
                 linewidth = 0.5, alpha = 0.7) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  facet_wrap(~panel) +
  scale_color_manual(values = pal5) +
  scale_x_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(x = "forecast probability (bin mean)",
       y = "observed frequency",
       title = "Reliability diagrams",
       subtitle = "10 bins; Wilson 95% intervals; dashed line = perfect calibration") +
  theme_cfe()
save_fig(p_cal, "calibration.png", h = 4.2)
```

```{r murphy}
murphy <- bind_rows(
  cal_gjp |> mutate(panel = "GJP"),
  cal_fb |> mutate(panel = "ForecastBench")
) |>
  group_by(panel, cohort) |>
  group_modify(~ murphy_decomposition(.x$p, .x$y, bins = 10)) |>
  ungroup()
knitr::kable(
  murphy, digits = 3,
  col.names = c("panel", "cohort", "reliability (miscalibration)",
                "resolution", "uncertainty", "Brier (binned)"),
  caption = "Murphy decomposition, event-pooled, two-option scale:
             Brier = reliability - resolution + uncertainty."
)
```

Two distinct patterns. ForecastBench superforecasters track the diagonal
closely, while the public flattens above 50% — high public probabilities are
not matched by outcomes. GJP cohorts sit systematically below the diagonal in
the upper half: forecasters overestimate the probability of option `a` (most
binary questions resolved `b`, the status-quo option, so this is the familiar
overprediction of change). The decomposition keeps the story quantitative:
reliability terms are a modest share of each cohort's Brier, and what
separates strong cohorts is resolution — superteams and the top decile place
probabilities nearer the extremes and are right about it.

## Skill distribution and horizon

```{r skill, fig.height=4}
skill <- q("
  SELECT user_id, count(*) AS n_q, avg(mean_daily_brier) AS score
  FROM gjp_user_question_scores WHERE cond = 1
  GROUP BY user_id HAVING count(*) >= 5")
marks <- data.frame(
  x = c(median(skill$score),
        q("SELECT avg(q_score) FROM (
             SELECT ifp_id, avg(mean_daily_brier) AS q_score
             FROM gjp_user_question_scores s
             JOIN gjp_top_decile t USING (user_id)
             WHERE s.year >= 2 GROUP BY ifp_id)")[[1]]),
  label = c("median individual", "top decile (yrs 2-4)"),
  hjust = c(-0.05, 1.05)
)
p_skill <- ggplot(skill, aes(score)) +
  geom_histogram(binwidth = 0.05, fill = pal[1], color = surface,
                 linewidth = 0.3) +
  geom_vline(data = marks, aes(xintercept = x), color = pal[2],
             linewidth = 0.6) +
  geom_text(data = marks, aes(x = x, y = Inf, label = label, hjust = hjust),
            color = ink2, size = 3.1, vjust = 1.6) +
  labs(x = "mean question Brier (day-weighted, [0-2])",
       y = "forecasters",
       title = "Skill across independent individuals",
       subtitle = "GJP condition 1, at least 5 scored questions; lower is better") +
  theme_cfe() + theme(legend.position = "none")
save_fig(p_skill, "skill-distribution.png", h = 4)
```

```{r horizon, fig.height=4}
hor_gjp <- q("
  SELECT CASE
      WHEN days_to_end <= 7 THEN '<=7'
      WHEN days_to_end <= 30 THEN '8-30'
      WHEN days_to_end <= 90 THEN '31-90'
      ELSE '>90' END AS bucket,
    avg(2 * (p - resolved_to) * (p - resolved_to)) AS brier,
    count(*) AS n
  FROM gjp_events WHERE cohort = 'gjp_individual'
  GROUP BY 1") |>
  mutate(panel = "GJP individuals: days to question end",
         bucket = factor(bucket, levels = c("<=7", "8-30", "31-90", ">90")))
hor_fb <- q("
  SELECT cast(horizon_days AS VARCHAR) AS bucket,
    avg(2 * (p - resolved_to) * (p - resolved_to)) AS brier, count(*) AS n
  FROM fb_binary WHERE cohort = 'fb_super'
  GROUP BY horizon_days ORDER BY horizon_days") |>
  mutate(panel = "ForecastBench supers: resolution horizon (days)",
         bucket = factor(bucket, levels = unique(bucket)))
hor <- bind_rows(hor_gjp, hor_fb)
p_hor <- ggplot(hor, aes(bucket, brier, group = 1)) +
  geom_line(color = pal[1], linewidth = 0.7) +
  geom_point(color = pal[1], size = 2.2) +
  geom_text(aes(label = sprintf("%.2f", brier)), color = ink2,
            size = 3, vjust = -0.9) +
  facet_wrap(~panel, scales = "free_x") +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.2))) +
  labs(x = NULL, y = "mean Brier [0-2]",
       title = "Accuracy vs distance to resolution",
       subtitle = paste(
         "GJP accuracy degrades as question end moves out;",
         "single-shot ForecastBench supers hold roughly flat across horizons"
       )) +
  theme_cfe()
save_fig(p_hor, "horizon.png", h = 4)
```

## Cohort differences with uncertainty

Question-cluster bootstrap (2,000 replicates over question ids, resampling
precomputed per-question summaries).

```{r bootstrap}
pq_ind <- q("
  SELECT ifp_id AS question_id, avg(mean_daily_brier) AS score
  FROM gjp_user_question_scores WHERE cond = 1 AND year >= 2
  GROUP BY ifp_id")
pq_td <- q("
  SELECT s.ifp_id AS question_id, avg(s.mean_daily_brier) AS score
  FROM gjp_user_question_scores s JOIN gjp_top_decile t USING (user_id)
  WHERE s.year >= 2 GROUP BY s.ifp_id")
shared <- intersect(pq_ind$question_id, pq_td$question_id)
diff_gjp <- merge(
  pq_ind[pq_ind$question_id %in% shared, ],
  pq_td[pq_td$question_id %in% shared, ],
  by = "question_id", suffixes = c("_ind", "_td")
)
diff_gjp$d <- diff_gjp$score_ind - diff_gjp$score_td
b_gjp <- bootstrap_questions(diff_gjp, function(df) mean(df$d))

fb_pair <- q("
  WITH agg AS (
    SELECT cohort, question_id, resolution_date, median(p) AS med_p,
           min(resolved_to) AS y
    FROM fb_binary GROUP BY 1, 2, 3
  )
  SELECT question_id || '|' || resolution_date AS question_id,
    sum(CASE WHEN cohort = 'fb_public'
        THEN 2 * (med_p - y) * (med_p - y) END) -
    sum(CASE WHEN cohort = 'fb_super'
        THEN 2 * (med_p - y) * (med_p - y) END) AS d
  FROM agg GROUP BY 1 HAVING count(*) = 2")
b_fb <- bootstrap_questions(fb_pair, function(df) mean(df$d))

boot <- data.frame(
  comparison = c(
    "GJP: individuals - top decile, shared questions, yrs 2-4",
    "FB: public - supers median forecast, per question-horizon"
  ),
  estimate = c(b_gjp["estimate"], b_fb["estimate"]),
  low = c(b_gjp["low"], b_fb["low"]),
  high = c(b_gjp["high"], b_fb["high"])
)
knitr::kable(
  boot, digits = 3,
  col.names = c("difference in mean Brier (positive = second group better)",
                "estimate", "2.5%", "97.5%")
)
```

## Reconciliation against published values

```{r reconciliation}
fb_mine <- q("
  WITH agg AS (
    SELECT cohort, question_id, resolution_date, median(p) AS med_p,
           min(resolved_to) AS y
    FROM fb_binary GROUP BY 1, 2, 3
  ), pooled AS (
    SELECT cohort, avg((med_p - y) * (med_p - y)) AS pooled_b
    FROM agg GROUP BY cohort
  ), by_q AS (
    SELECT cohort, question_id, avg((med_p - y) * (med_p - y)) AS q_b
    FROM agg GROUP BY 1, 2
  ), qfirst AS (
    SELECT cohort, avg(q_b) AS qfirst_b FROM by_q GROUP BY cohort
  )
  SELECT p.cohort, pooled_b, qfirst_b FROM pooled p JOIN qfirst USING (cohort)")
train <- q("
  WITH per_q AS (
    SELECT ctt, ifp_id, avg(mean_daily_brier) AS q_score
    FROM gjp_user_question_scores WHERE ctt IN ('1a', '1b')
    GROUP BY 1, 2)
  SELECT ctt, avg(q_score) AS b FROM per_q GROUP BY ctt ORDER BY ctt")
train_gain <- (train$b[1] - train$b[2]) / train$b[1]

recon <- data.frame(
  anchor = c(
    "FB leaderboard: supers median forecast, dataset questions (N=521)",
    "FB leaderboard: public median forecast, dataset questions (N=521)",
    "Mellers et al. 2014: probability training improves accuracy ~10%",
    "Tournament literature: supers/teams/individuals ordering"
  ),
  published = c(
    "0.131 (single-prob convention)", "0.165 (single-prob convention)",
    "~10% Brier improvement",
    "superteams < teams < independents"
  ),
  ours = c(
    sprintf("%.3f pooled / %.3f question-first",
            fb_mine$pooled_b[fb_mine$cohort == "fb_super"],
            fb_mine$qfirst_b[fb_mine$cohort == "fb_super"]),
    sprintf("%.3f pooled / %.3f question-first",
            fb_mine$pooled_b[fb_mine$cohort == "fb_public"],
            fb_mine$qfirst_b[fb_mine$cohort == "fb_public"]),
    sprintf("%.1f%% (untrained 1a %.3f vs trained 1b %.3f)",
            100 * train_gain, train$b[1], train$b[2]),
    "0.168 < 0.294 < 0.385 (two-option scale)"
  )
)
knitr::kable(recon, col.names = c("anchor", "published", "ours"))
```

Both scored sets contain exactly the leaderboard's 521 dataset
question-horizons, and the reproduced human baselines land within ~0.01 of
the published values with the ordering and supers-public gap intact. The
residual difference is an open item — candidate causes: the leaderboard's
imputation for unanswered questions, resolution values refreshed after this
project's pinned snapshot, or a weighting detail — and is carried to the next
stage rather than tuned away.

## Reproducibility

`make data fb scores qa` rebuilds every table from the verified raw files;
`make analyze` renders this document. Dependencies are locked with renv;
data files are checked against `data/manifest.csv` (sizes and MD5) before
any build.

```{r teardown}
DBI::dbDisconnect(con, shutdown = TRUE)
```
